---
title: "CVD MR"
output: html_document
date: "2024-02-19"
editor_options: 
  markdown: 
    wrap: 72
---

Full MR script for one exposure and multiple outcomes.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	comment = NULL
)
#Try MR
source("setup_Arja.R")

# required columns for outcome: rsid, beta, se, eaf, effect_allele_col, other_allele_col, pval
```

The next two chunks are for reading in the exposure data and the outcome
data. Add your own file names.

```{r eval = FALSE}
# read in exposure data  -  glucose
exposure_data = read_exposure_data(
  filename = here("/Users/angelakoloi/Desktop/yfs metabolites/2007/data/MR analysis/serum albuminRSID.tsv"),
  sep = "\t",
  snp_col = "RSID",
  beta_col = "beta", 
  se_col = "standard_error",
  eaf_col = "effect_allele_frequency",
    effect_allele_col = "effect_allele",
    other_allele_col = "other_allele",
  log_pval = FALSE,
  pval_col = "p_value",
  chr_col = "CHR",
  pos_col = "BP"
)
exposure_data$exposure <- "serum albumin"
```

```{r eval = FALSE}
cvd_traits <- c("PGC_UKB_depression", "ALLSTROKE", "IS", "CES", "LAS", "SVD", "CAD", "CAC", "CIMT" ) 

outcome_list <- list()

for (i in cvd_traits) {
  
  # read in outcome data, CVD
  file <- paste0(i, "_with_RSID.tsv")
  cat(file) 
  
  outcome_data = read_outcome_data(
    filename = here("/Users/angelakoloi/Desktop/yfs metabolites/2007/data/MR analysis/", file),
    sep = "\t",
    snps = exposure_data$SNP, 
    snp_col = "RSID",
    beta_col = "beta",#calculate beta for the exposure 
    se_col = "standard_error",
    eaf_col = "effect_allele_frequency",
    effect_allele_col = "effect_allele",
    other_allele_col = "other_allele",
    pval_col = "p_value",
    #ncase_col = "NCAS",
    #ncontrol_col = "NCON",
    log_pval = FALSE,
    chr_col = "CHR",
    pos_col = "BP"
  )
  outcome_data$outcome <- i
  
  outcome_list[[i]] <- outcome_data
  
}

outcome_list
#save(outcome_list, file = here("/Users/angelakoloi/Desktop/yfs metabolites/2007/data/MR analysis/", "outcome_forMR_serum albumin.RData")) 
```

Preprocess the data: The exposure data is filtered and clumped to
identify instruments. Then one harmonised dataframe is created from
exposure and outcome.

```{r preprocessing, eval = FALSE}
# exposure is always from exposure_data, but outcome_data comes from outcome_list

SNP_numbers <- c() # store how many SNPs are left after each filtering step
preprocessed_list <- list()

for ( i in cvd_traits){
  outcome_data <- outcome_list[[i]]
  
  # Keep only SNPs that are in both datasets
  # !! careful: refresh outcome data on every iteration but keep original exposure_data object the same
  exposure_data_small <- exposure_data[exposure_data$SNP %in% outcome_data$SNP,]
  outcome_data <- outcome_data[outcome_data$SNP %in% exposure_data_small$SNP,]
  
  # Filtering and clumping in the exposure data
  # pvalue filter
  exposure_data_topsnps = exposure_data_small[exposure_data_small$pval.exposure < 5e-08, ]
  # maf filter
  exposure_data_topsnps_maf = exposure_data_topsnps %>% filter(eaf.exposure < 0.99 & eaf.exposure > 0.01)
  exposure_data_clumped <- clump_data(exposure_data_topsnps_maf)
  
  
  # clumping
  # using own function wrappers of plink clumping to adapt to using local ld files rather than server
  # plink_bin <- "C:/Users/emma/AppData/Local/R/win-library/4.3/genetics.binaRies/bin/plink2.exe"
  # bfile = here("ld_files", "g1000_eur")
  #exposure_data_clumped <- clump_data_locally( # defaults are clump_kb = 10000, clump_r2 = 0.001
   # dat = exposure_data_topsnps_maf, 
   # bfile = "/Users/angelakoloi/Desktop/yfs metabolites/2007/data/MR analysis/g1000_eur",
   # plink_bin = "/Users/angelakoloi/Downloads/plink_mac_20231211")
 
  # for server version use the TwoSampleMR version clump_data(). only required input argument is dat = your dataset

  harmonised_data <- harmonise_data(exposure_dat = exposure_data_clumped, outcome_dat = outcome_data)
  
  # Store some info about filtering process
  SNP_numbers <- c()
  SNP_numbers["beforefilter"] <- dim(exposure_data_small)[1]
  SNP_numbers["pval"] <- dim(exposure_data_topsnps)[1]
  SNP_numbers["maf"] <- dim(exposure_data_topsnps_maf)[1]
  SNP_numbers["clumped"] <- dim(exposure_data_clumped)[1]
  SNP_numbers["harmonised"] <- dim(harmonised_data)[1]
  
  # And return something
  preprocessed_list[[i]] <- list(data = harmonised_data, log = SNP_numbers)
}

preprocessed_list
#save(preprocessed_list, file = here("/Users/angelakoloi/Desktop/yfs metabolites/2007/data/MR analysis/", "all_processed_serum albumin.RData"))
```

Run the main MR analysis and also all the essential follow-up analyses.

As described by Yuri:

"First, a series of univariable 2SMR analyses were performed based on
the inverse variance weighted (IVW) estimator (Burgess et al., 2013),
pooling SNP-exposure/SNP-outcome estimates inversely weighted by their
standard error. Since IVW assumes that all SNPs are valid instruments or
that the sum of the directional bias is zero, the robustness of
significant results was tested in sensitivity analyses based on weighted
median and MR-Egger estimators. The weighted median (Bowden et al.,
2016) estimator is the median of the weighted empirical distribution
function of individual SNP ratio estimates, providing consistent effect
estimates even if half of the instruments are invalid. The MR-Egger
regression (Bowden et al., 2015) consists of a weighted linear
regression similar to IVW relying on the InSIDE assumption (the
magnitude of any pleiotropic effects should not correlate with the
magnitude of the main effect), providing valid effect estimate even if
all SNPs are invalid instruments under the 'NOME' assumption
(uncertainty in the SNPexposure association estimates is negligible)
(Bowden et al., 2017). At least 10 genetic instruments are recommended
(Bowden et al., 2015) to run adequately powered MR-Egger analyses.
Furthermore, heterogeneity among included SNPs was tested via Cochran's
Q test, single SNP, and leave-one-out SNP analyses. The presence of
potential horizontal pleiotropy (a genetic instrument for exposure
influencing the outcome by mechanisms other than exposure) was tested
using the MR-Egger intercept (Bowden et al., 2017) and the MR-PRESSO
(pleiotropy residual sum and outlier) method (Verbanck et al., 2018)
(supplemental methods). Finally, we performed reversed univariable 2SMR
analyses testing the potential causal impact of depression liability on
AC circulating levels."

```{r results, eval = FALSE}
#load(here("sumstats", "all_processed_serum albumin.RData"))
main_results_mr <- list()

for (i in names(preprocessed_list)) {
  preprocessed_data <- preprocessed_list[[i]][["data"]]
  
  if (nrow(preprocessed_data) < 2) {
    cat("Only one SNP for", i, "\n")
    next
  }
  
  # Run MR using various methods
  mr_output <- mr(preprocessed_data, method_list = c("mr_ivw", "mr_weighted_median", "mr_egger_regression"))
  mr_or <- generate_odds_ratios(mr_output)
  
  # Print MR results
  print(mr_or)
  
  # Scatter plot
  scatterplot <- mr_scatter_plot(mr_output, preprocessed_data)
  
  # Instrument strength
  fstatistic <- (preprocessed_data$beta.exposure^2) / (preprocessed_data$se.exposure^2)
  f_range <- c(min(fstatistic), max(fstatistic))
  
  # Initialize lists for optional results
  mr_het <- mr_leave1out <- oneout_plot <- mr_1snp <- onein_plot <- mr_pleiot <- mr_presso_output <- NULL
  
  # Conditional checks for SNP count
  if (nrow(preprocessed_data) > 1) {
    # Heterogeneity
    if (length(unique(preprocessed_data$SNP)) > 1) {
      mr_het <- tryCatch(mr_heterogeneity(preprocessed_data), error = function(e) NULL)
      mr_leave1out <- tryCatch(mr_leaveoneout(preprocessed_data), error = function(e) NULL)
      oneout_plot <- tryCatch(mr_leaveoneout_plot(mr_leave1out), error = function(e) NULL)
      mr_1snp <- tryCatch(mr_singlesnp(preprocessed_data), error = function(e) NULL)
      onein_plot <- tryCatch(mr_forest_plot(mr_1snp), error = function(e) NULL)
    } else {
      cat("Not enough SNPs available for heterogeneity analysis of '", i, "'\n", sep = "")
    }
    
    # Horizontal Pleiotropy
    if (length(unique(preprocessed_data$SNP)) > 2) {
      mr_pleiot <- tryCatch(mr_pleiotropy_test(preprocessed_data), error = function(e) NULL)
      mr_presso_output <- tryCatch(
        run_mr_presso(preprocessed_data, NbDistribution = 1000, SignifThreshold = 0.05),
        error = function(e) NULL
      )
    } else {
      cat("Not enough SNPs available for pleiotropy analysis of '", i, "'\n", sep = "")
    }
  } else {
    cat("Not enough SNPs available for analysis of '", i, "'\n", sep = "")
  }
  
  # Print results for pleiotropy test
  if (!is.null(mr_pleiot)) print(mr_pleiot)
  
  # Store results in list
  main_results_mr[[i]] <- list(
    mr_output = mr_output, mr_or = mr_or, scatterplot = scatterplot,
    f_range = f_range, mr_het = mr_het, mr_leave1out = mr_leave1out,
    oneout_plot = oneout_plot, mr_1snp = mr_1snp, onein_plot = onein_plot,
    mr_pleiot = mr_pleiot, mr_presso_output = mr_presso_output
  )
}

main_results_mr

#save(main_results_mr, file = here("/Users/angelakoloi/Desktop/yfs metabolites/2007/data/MR analysis/Results", "MR_DEP-CVD_results_serum albumin.RData"))
```

# Main results (basic OR)

```{r}
#load(here("Results", "MR", "MR_DEP-CVD_results_serum albumin.RData"))

# Nice table of main results
or_full_df <- as.data.frame(do.call("rbind", map(main_results_mr, "mr_or")))[c(3,4,5,6,12,13,14,9)]
or_full_df$pval <- formatC(or_full_df$pval, format = "e", digits = 2) # pvalues to scientific notation
or_full_df <- or_full_df %>% 
  rename("or.lci95" = "or_lci95", "or.uci95" = "or_uci95") %>% # change separator
  mutate(across(c(or, or.lci95, or.uci95), ~ round(.x, digits = 2))) # round to two digits

or_wide_df <- pivot_wider(
  data = or_full_df,
  names_from = method,
  names_vary = "slowest",
  names_glue = "{method}_{.value}",
  values_from = c(or, or.lci95, or.uci95, pval)
)

flextable(or_wide_df) %>% separate_header(split = "_") %>%  fontsize(size = 14) %>% fit_to_width(8) %>% autofit()
```

Note that we didn't expect an effect for CES (mostly genes specific to
heart development) and indeed the estimate is not significant.

```{r}
# load(here("Results", "CVD_mr_results_serum albumin.RData"))
# 
# # Nice table of main results
full_df <- as.data.frame(do.call("rbind", map(main_results_mr, "mr_output")))[c(3,4,5,6,7,8,9)]
or_full_df$pval <- formatC(or_full_df$pval, format = "e", digits = 2) # pvalues to scientific notation
or_full_df <- or_full_df %>% 
rename("or.lci95" = "or_lci95", "or.uci95" = "or_uci95") %>% # change separator
mutate(across(c(or, or.lci95, or.uci95), ~ round(.x, digits = 2))) # round to two digits
# 
or_wide_df <- pivot_wider(
   data = or_full_df,
   names_from = method,
   names_vary = "slowest",
   names_glue = "{method}_{.value}",
   values_from = c(or, or.lci95, or.uci95, pval)
)
# 
flextable(or_wide_df) %>% separate_header(split = "_")
main_results_mr
```

```{r fig.dim = c(8, 17)}
# Plot collection
# plots <- unlist(map(main_results_mr, "scatterplot"))
p1 <- main_results_mr$ALLSTROKE$scatterplot$v3kGW8.7mNsA8
p2 <- main_results_mr$IS$scatterplot$v3kGW8.gYxmey
p6 <- main_results_mr$CAD$scatterplot$v3kGW8.AEKi7w
p4 <- main_results_mr$CES$scatterplot$v3kGW8.a4pcPh
p5 <- main_results_mr$LAS$scatterplot$v3kGW8.bpulR3
p3 <- main_results_mr$SVD$scatterplot$v3kGW8.KeIQy5
p7 <- main_results_mr$CAC$scatterplot$e7VlWv.JiGy5p
p8 <- main_results_mr$CIMT$scatterplot$e7VlWv.fv9RyK

p1 + p2 + p3 + p4 + p5 + p6 + p7 + p8 +
  plot_layout(ncol = 2, guides = "collect") &
  theme(legend.position = "bottom")
```

# MR more analyses

## Instrument strength

```{r }
# # Test instrument strength
# fstatistic <- (harmonised_data$beta.exposure^2) / (harmonised_data$se.exposure^2) # approximation for sumstats
# min(fstatistic) # F should not be < 10 (Pierce et al. 2011)
# max(fstatistic)

fstats <- as.data.frame(do.call("rbind", map(main_results_mr, "f_range")))
fstats <- cbind(or_wide_df$nsnp, round(fstats, 4))
colnames(fstats) <- c("N SNPs", "F-statistics_min", "F-statistics_max" )

flextable(fstats %>% rownames_to_column( var = "Outcome")) %>% separate_header()
```

Probably these make sense..? There is large overlap in the SNPs involved
for each outcome, so the SNP with the weakest and the SNP with the
strongest effect on the exposure could feasibly be the same for all
stroke outcomes.

# Heterogeneity

## Cochran's Q

A significant Cochran's Q indicates presence of an outlier.

```{r }
# # Heterogeneity
# mr_het <- mr_heterogeneity(harmonised_data) # this give Cochran's Q (if significant indicates outlier), one for Egger + one for IVW
qstats <- as.data.frame(do.call("rbind", map(main_results_mr, "mr_het")))
qstats["Q"] <- round(qstats["Q"])
qstats["df"] <- qstats["Q_df"]
qstats["pvalue"] <- formatC(qstats$Q_pval, format = "e", digits = 2)
qstats <- cbind(qstats[c("outcome", "exposure", "method", "Q", "df", "pvalue")])

q_wide <- pivot_wider(
  data = qstats,
  names_from = method,
  names_vary = "slowest",
  names_glue = "{method}_{.value}",
  values_from = c(Q, df, pvalue)
)

flextable(q_wide) %>% separate_header(split = "_") %>% add_header_row(values = c("", "Cochran's Q"), colwidths = c(2, 6))
```

## Leave one out analysis

```{r }
map(main_results_mr, "oneout_plot")
```

## Single SNP analysis

```{r }
map(main_results_mr, "onein_plot")
```

# Horizontal Pleiotropy

## MR-Egger Intercept

```{r }
# Pleiotropy
mr_pleiot <- as.data.frame(do.call("rbind", map(main_results_mr, "mr_pleiot")))
mr_pleiot[c("egger_intercept", "se")] <- round(mr_pleiot[c("egger_intercept", "se")], 5)
mr_pleiot["pval"] <- formatC(mr_pleiot$pval, format = "e", digits = 2)

flextable(mr_pleiot[-c(1,2)])
```

## MR presso

```{r}
# MR_presso
mr_presso <- map(main_results_mr, c("mr_presso_output"))

outlier_corrected <- do.call( "rbind", 
                              lapply(mr_presso, function(x) {
                                x[[1]][["Main MR results"]][2, c("Causal Estimate", "Sd", "P-value")]
                                }))

global_test <- as.data.frame(do.call( "rbind", 
                              lapply(mr_presso, function(x) {
                                unlist(x[[1]][["MR-PRESSO results"]][["Global Test"]])
                                })))

distortion_coefficient <- as.data.frame(do.call( "rbind", 
                              lapply(mr_presso, function(x) {
                                x[[1]][["MR-PRESSO results"]][["Distortion Test"]][["Distortion Coefficient"]]
                                })))
distortion_pvalue <- as.data.frame(do.call( "rbind", 
                              lapply(mr_presso, function(x) {
                                unlist(x[[1]][["MR-PRESSO results"]][["Distortion Test"]][["Pvalue"]])
                                })))


mr_presso_table <- cbind(or_wide_df$exposure, or_wide_df$nsnp, global_test, outlier_corrected, distortion_coefficient, distortion_pvalue)
colnames(mr_presso_table) <- c("Exposure", "N SNPs", "GlobalTest_RSS", "GlobalTest_pvalue", "OutlierCorrected_estimate", "OutlierCorrected_se", "OutlierCorrected_pvalue", "DistortionTest_Coefficient", "DistortionTest_pvalue")
mr_presso_table$OutlierCorrected_pvalue <- formatC(mr_presso_table$OutlierCorrected_pvalue, format = "e", digits = 2)
mr_presso_table$GlobalTest_RSS <- round(as.numeric(mr_presso_table$GlobalTest_RSS), 3)


flextable(mr_presso_table %>% rownames_to_column( var = "Outcome")) %>% separate_header() %>% fit_to_width(8)
```
